Motional Ground-State Cooling Outside the Lamb-Dicke Regime -- Supplemental material

### 2017-08-14 paper.tex
SM update and other fixup

\begin{figure*}
  \includegraphics[height=4.5cm]{fig1_combined.pdf}
  \caption{Single Na atom Raman sideband cooling scheme and sequence. (A)
    The Raman transitions between $|2,-2;n\rangle$ and $|1,-1;n+\Delta n\rangle$ have a one-photon detuning $\Delta=25$ GHz below the $3^2S_{1/2}$ to $3^2P_{3/2}$ transition. Two-photon detuning, $\delta$, is defined relative to the $\Delta n=0$ carrier transition. For optical pumping, we use two $\sigma^-$ polarized transitions, one to pump the atom state out of $|1,-1\rangle$ via $3^2P_{3/2}$ and one to pump atoms out of $|2,-1\rangle$ via $3^2P_{1/2}$
     to minimize heating of the $|2,-2\rangle$ state.
    (B) Geometry and polarizations of the Raman and optical pumping beams relative to the
    optical tweezer and bias magnetic field.  Raman beams R1 and R4 address the radial $x$-mode. R1 and R2 address the radial $y$-mode.  R3 and R4 address the axial $z$-mode, where the beams also couple to radial motion, but this coupling can be neglected when the atoms is cooled to the ground state of motion.  %, which also couples to the radial $y$ mode.    The radial $x$-mode is addressed by turning on beams R1 and R4, $y$ is addressed by R1 and R2, and $z$ is addressed by R3 and R4.  In this geometry, the radial mode directions coincide with the respective $k$-vector differences.  The ``axial'' Raman beams also couple to radial motion, but this coupling can be neglected when the atoms is cooled to the ground state of motion.
    (C) Schematic of the cooling pulse sequence. The tweezer is strobed at 3 MHz to
    reduce light shifts during optical pumping~\cite{Hutzler2017-LightShifts}.
    Each cooling cycle consists of $8$ sideband pulses.
    The four axial pulses address two sideband orders.
    The two pulses in each radial direction either address $\Delta n=-2$ and $\Delta n=-1$
    or have different durations to drive $\Delta n=-1$, at the end of the cooling sequence when most of the population is below $n=3$.
    The Raman cooling and spectroscopy pulses have Blackman envelopes to reduce off-resonant
    coupling, while the measurement Rabi pulses in Fig.~\ref{f-radial}(B,C) and \ref{f-axial}(B,C) have square envelopes to simplify analysis.
    \label{f-setup}}
\end{figure*}

Specifically, Raman transitions are driven in $^{23}$Na between the hyperfine states
$|2, -2\rangle$ and $|1, -1\rangle$ in the presence of an $8.8$ G magnetic field. The external field is
orthogonal to the effective magnetic field of the tweezer, to reduce vector light shifts~\cite{Kaufman2012,Thompson2013}.
Subsequently, an OP pulse, which consists of two frequencies both with $\sigma^-$-polarization, is applied to bring the atom out of the $F=1$ manifold and back to $|2, -2\rangle$ via spontaneous emission (Fig.~\ref{f-setup}A).
%brings the atom back to $|F=2, m_F=-2\rangle$ via spontaneous emission.
%, which consists of two frequencies with $\sigma^-$-polarization for the to bring the atom out of $F=1$ manifold and back to $|F=2, m_F=-2\rangle$ via spontaneous emission.
It is important that the optical pumping step scatter as few photons as possible to minimize recoil heating. %The OP consists of driving a $|F=1, m_F=-1\rangle$ into the $F=2$ manifold via the transition of $3P_{3/2} D2 line$
Therefore, to optically pump atoms from $|2,-1\rangle$ into $|2, -2\rangle$ while keeping the target state dark, we apply light resonant with $3^2P_{1/2}$~\cite{Monroe1995, Grobner2017}~\footnote{We find a reduction in the scattering rate by a factor of $130(20)$, as compared to using an OP resonant with $3^2P_{3/2}$, from which the $|2, -2>$ state could always scatter a photon via the excited $|F'=3, m_F'=-3>$ state.}.
%~\footnote{We find a reduction in the scattering rate by a factor of $130(20)$, as compared to using an OP resonant with $3P_{3/2}$, from which the $|2, -2 \rangle$}

>>>>

\begin{figure*}
  \includegraphics[height=4.5cm]{fig1_combined.pdf}
  \caption{Single Na atom Raman sideband cooling scheme and sequence. (A)
    The Raman transitions between $|2,-2;n\rangle$ and $|1,-1;n+\Delta n\rangle$ have a one-photon detuning $\Delta=25$ GHz below the $3^2S_{1/2}$ to $3^2P_{3/2}$ transition. Two-photon detuning, $\delta$, is defined relative to the $\Delta n=0$ carrier transition. For optical pumping, we use two $\sigma^-$ polarized transitions, one to pump the atom state out of $|1,-1\rangle$ via $3^2P_{3/2}$ and one to pump atoms out of $|2,-1\rangle$ via $3^2P_{1/2}$
     to minimize heating of the $|2,-2\rangle$ state.
    (B) Geometry and polarizations of the Raman and optical pumping beams relative to the
    optical tweezer and bias magnetic field.  Raman beams R1 and R4 address the radial $x$-mode. R1 and R2 address the radial $y$-mode.  R3 and R4 address the axial $z$-mode, where the beams also couple to radial motion, but this coupling can be neglected when the atoms is cooled to the ground state of motion.  %, which also couples to the radial $y$ mode.    The radial $x$-mode is addressed by turning on beams R1 and R4, $y$ is addressed by R1 and R2, and $z$ is addressed by R3 and R4.  In this geometry, the radial mode directions coincide with the respective $k$-vector differences.  The ``axial'' Raman beams also couple to radial motion, but this coupling can be neglected when the atoms is cooled to the ground state of motion.
    (C) Schematic of the cooling pulse sequence. The tweezer is strobed at 3 MHz to
    reduce light shifts during optical pumping~\cite{Hutzler2017-LightShifts}.
    Each cooling cycle consists of $8$ sideband pulses.
    The four axial pulses address two sideband orders.
    The two pulses in each radial direction either address $\Delta n=-2$ and $\Delta n=-1$
    or have different durations to drive $\Delta n=-1$, at the end of the cooling sequence when most of the population is below $n=3$.
    The Raman cooling and spectroscopy pulses have Blackman envelopes~\cite{Kasevich1992}
    to reduce off-resonant coupling,
    while the measurement Rabi pulses in Fig.~\ref{f-radial}(B,C) and \ref{f-axial}(B,C)
    have square envelopes to simplify analysis.
    \label{f-setup}}
\end{figure*}

Specifically, Raman transitions are driven in $^{23}$Na between the hyperfine states
$|2, -2\rangle$ and $|1, -1\rangle$ in the presence of an $8.8$ G magnetic field. The external field is
orthogonal to the effective magnetic field of the tweezer, to reduce vector light shifts~\cite{Kaufman2012,Thompson2013}.
Subsequently, an OP pulse, which consists of two frequencies both with $\sigma^-$-polarization, is applied to bring the atom out of the $F=1$ manifold and back to $|2, -2\rangle$ via spontaneous emission (Fig.~\ref{f-setup}A).
%brings the atom back to $|F=2, m_F=-2\rangle$ via spontaneous emission.
%, which consists of two frequencies with $\sigma^-$-polarization for the to bring the atom out of $F=1$ manifold and back to $|F=2, m_F=-2\rangle$ via spontaneous emission.
It is important that the optical pumping step scatter as few photons as possible to minimize recoil heating. %The OP consists of driving a $|F=1, m_F=-1\rangle$ into the $F=2$ manifold via the transition of $3P_{3/2} D2 line$
Therefore, to optically pump atoms from $|2,-1\rangle$ into $|2, -2\rangle$ while keeping the target state dark, we apply light resonant with $3^2P_{1/2}$~\cite{Monroe1995, Grobner2017}~\footnote{We find a reduction in the scattering rate by a factor of $130(20)$, as compared to using an OP resonant with $3^2P_{3/2}$, from which the $|2, -2>$ state could always scatter a photon via the excited $|F'=3, m_{F'}=-3>$ state.}.
%~\footnote{We find a reduction in the scattering rate by a factor of $130(20)$, as compared to using an OP resonant with $3P_{3/2}$, from which the $|2, -2 \rangle$}

### 2017-08-14 sm.tex
[sm] Update author list
Kang-Kuen's edit and white space fix
SM update and other fixup

\begin{document}
\title{Motional Quantum Ground-State Cooling Outside the Lamb-Dicke Regime -- Supplemental material}
\author{Yichao Yu}
\email{yichaoyu@g.harvard.edu}
\author{Nicholas R. Hutzler}
\altaffiliation{Present address: California Institute of Technology, Division of Physics, Mathematics, and Astronomy. Pasadena, CA, 91125}
\author{Jessie T. Zhang}
\author{Lee R. Liu}
\author{Kang-Kuen Ni}
\email{ni@chemistry.harvard.edu}
\affiliation{Department of Chemistry and Chemical Biology, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Department of Physics, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Harvard-MIT Center for Ultracold Atoms, Cambridge, Massachusetts, 02138, USA}

@@

\section{Simulation of 3D Raman sideband cooling}

We use a computer simulation to study the importance of different effects
and guide the optimization of Raman sideband cooling.
The simulation uses the quantum jump (or Monte-Carlo wavefunction) method~\cite{Dalibard1992,Chretien2014}
to accurately calculate the evolution within each laser pulse where different energy eigenstates
are assumed to decohere immediately upon spontaneous emission and on pulse boundaries.
Such approximations are valid and greatly simplify the computation
at the cost of limiting the applicability of the simulation to processes that require stronger
coherence, such as Ramsey spectroscopy or rapid push-out of the atoms from the trap
(faster than a trap oscillation period).

A number of other simplifying approximations, listed below, are also used in the simulation.
These either have a demonstrably negligible effect on the experiment,
or result in a significant reduction of complexity for an effect
that can be optimized more easily in the experiment.

\begin{enumerate}
\item \emph{All Raman pulses resonantly drive only one sideband/carrier order.}  We tweak the pulse shape, power and length as discussed in the text, for example by using Blackman pulses, to reduce coupling to other orders. We do see deviation from this approximation in the experiment, though mostly in the long detection pulses, and it is less of a concern for the short cooling pulses.
\item \emph{No heating from other sources.}\\
  We measured the heating rate and found that it results mainly from off-resonant scattering
  of photons, which is included in the simulation.  In particular, we can ignore heating from the switching trap.
\item \emph{Same trapping frequency for all states.}\\
  This effect is negligible for Na in the ground state due to the small fine structure splitting in the excited state.
  It is straightforward to include this effect if the calculation needs to be repeated
  on a system where this is not the case (e.g. directly laser cooled molecules).
\end{enumerate}

Some effects that are important to include are all three dimensions (3D motional states,
accurate 3D wavevector, and different emission pattern for different photon polarizations),
off-resonant scattering from all laser beams, and optical pumping polarization imperfection.
The optical pumping polarization and off-resonant scattering are especially important
for accurate simulation since they are the main limitation on cooling performance.

In general, each pulse has a coherent part from the Raman drive
(which is absent for the optical pumping pulses)
and an incoherent part from scattering (optical pumping or off-resonant).
This nicely fits the general approach of the quantum jump method.
Due to the large number of states that we need to take into account
($30-150$ motional per axis and $8$ hyperfine, giving a total of $10^6$),
generic wavefunction propagation is very inefficient.
Therefore, we further simplify by reducing the number of states tracked during
the coherent propagation and applying quantum jumps using a set of reduced jump operators that
describe the total jump probability from a certain state (summed over all final states).
The coherent propagation is then accurately solved in order to avoid having to use a time step
much shorter than the inverse of scattering rate (up to hundreds of kHz)\footnote{\href{https://goo.gl/EZ13wt}{note about modification to the quantum jump method (https://goo.gl/EZ13wt)}}.

In order to simulate a pulse, we first start with the coherent propagation that includes the
modification from the jump operators as in the quantum jump method.
When a jump happens, we pick a final internal state to jump to based on the relative rates
of different scattering sources and their branching ratios.
This process determines the polarization and emission pattern of the emitted photon,
from which we randomly sample a particular wave vector.
The photon recoil being defined, we can then assign a final motional state to the atom.
This process is then repeated as long as the pulse is on,
before we start to simulate the next pulse.

We apply ``virtual'' measurements on the results of each simulated cooling sequence
which we then average to obtain the expected value of experimental quantities.
Due to the decoherence approximation, the measurement has to commute with the Hamiltonian.
Particularly useful measurements used in our simulation are hyperfine state distribution,
average motional energy, loss rate (where an atom is assumed to be lost if its motional energy
goes above a certain threshold at any time) probability of being in certain states
(e.g. ground state).

\begin{figure}[t]
  \includegraphics[width=8.5cm]{imgs/simcool_no_scatter.pdf}
  \caption{Simulated 3D ground state population as a function of cooling cycles without
    technical imperfections. The ground state probability reaches $>99\%$. \label{f-no-scatter}}
\end{figure}
\begin{figure}[t]
  \includegraphics[width=8.5cm]{imgs/simcool_real.pdf}
  \caption{Simulated 3D ground state population as a function of cooling cycles including effect of
    off-resonant scattering from all beams and OP misalignment.
    For this particular sequence, the ground state probability saturates to $\approx88\%$.
    \label{f-real}}
\end{figure}

If off-resonant scattering and optical pumping polarization misalignment are ignored,
the simulation shows that there is not a cooling limit
despite the large Lamb-Dicke parameter and high initial temperature (Fig \ref{f-no-scatter}).
However, when these effects are taken into account, the ground state probability can saturate at
a lower value (Fig \ref{f-real}) and require much more careful tweaking.
The change of slop in these plots corresponds to changing the sideband orders used for cooling.

The simulation is written in Julia\cite{Bezanson2017}.
All code is available under the GPLv3 license at
\footnote{\href{https://goo.gl/WLJwJp}{Library code (https://goo.gl/WLJwJp)}} and
\footnote{\href{https://goo.gl/XXVreq}{Top level code (https://goo.gl/XXVreq)}}

\section{Cooling sequence}

\subsection{Optical pumping parameters}
All optical pumping pulses has the same length of $40\mu$s
with a $|F=1,m_F=-1\rangle$ scattering rate of $0.14$ MHz
and a $|F=2,m_F=-1\rangle$ scattering rate of $0.39$ MHz.

@@

that ramps with a Blackman profile during the pulse.
$\delta'$ is the two photon detuning relative to the zero field HF splitting

@@

than the zero field HF splitting.

>>>>

\begin{document}
\title{Motional Ground-State Cooling Outside the Lamb-Dicke Regime -- Supplemental material}
\author{Y. Yu}
\email{yichaoyu@g.harvard.edu}
\author{N. R. Hutzler}
\altaffiliation{Present address: California Institute of Technology, Division of Physics, Mathematics, and Astronomy. Pasadena, CA, 91125}
\author{J. T. Zhang}
\author{L. R. Liu}
\author{T. Rosenband}
\author{K.-K. Ni}
\email{ni@chemistry.harvard.edu}
\affiliation{Department of Chemistry and Chemical Biology, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Department of Physics, Harvard University, Cambridge, Massachusetts, 02138, USA}
\affiliation{Harvard-MIT Center for Ultracold Atoms, Cambridge, Massachusetts, 02138, USA}

@@

\section{Raman Sideband Cooling Pulse Sequence}

\subsection{Optical pumping parameters}
The optical pumping is done with a $\sigma^-$ polarized beam aligned with the bias magnetic field.
There are two frequencies in the beam that are on resonance with
the $|F=2,m_F=-1\rangle$ to $|F'=2,m_{F'}=-2\rangle$ D1 transition
and the $|F=1,m_F=-1\rangle$ to $|F'=2,m_{F'}=-2\rangle$ D2 transition respectively.
All optical pumping pulses have the same length of $40\mu$s
with a $|F=1,m_F=-1\rangle$ scattering rate of $0.14$ MHz
and a $|F=2,m_F=-1\rangle$ scattering rate of $0.39$ MHz.

@@

that ramps with a Blackman profile during the pulse~\cite{Kasevich1992}.
$\delta'$ is the two photon detuning relative to the zero field hyperfine splitting

@@

than the zero field hyperfine splitting.
